---
title: "Calculate Deriviate of GAMM Smoothed Line"
author: "Nathan Brouwer"
date: "January 30, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preliminaries

### Load packages

```{r}
library(nlme)    #required by mgcv
library(mgcv)    #for gam

library(ggplot2) #plotting
library(cowplot) #ggplot extenion
```


### Load data

#### Load raw data

Raw data has measurements every 30 minutes.

```{r}
par1 <- read.csv("./data/garden_sensor_data_cleaned.csv")
```


#### Load summarized data
Load garden data that has been summarized by mid-day mean for plotting exploratory graph.

```{r}
par.mean <- read.csv("./data/garden_sensor_data_10amTo2pm_mean.csv")
```



## Exploratory plots

### Plot raw data

```{r}
qplot(y = PAR,
        x = (julian),
          geom = c("point","smooth"),
          data = par1,
      color = hour)  +
  theme_bw() +
   xlab("Date")
```


### Plot daily means

We can fit a LOESS smoother using geom = "smooth" or stat_smooth().

```{r}
qplot(y = mean,
        x = (julian),
          geom = c("point","smooth"),
          data = par.mean) +
    geom_errorbar((aes(ymin = mean-2*SE, 
                       ymax = mean+2*SE))) +
  theme_bw() + ylab("Mid-day mean PPFD") +
   xlab("Date")
```

ggplot is very flexible and can actually call the mgcv library and fit a GAM on the fly.  Note that I removed "smooth" from the call to geom = within qplot and added stat_smooth() at the end.  The default for stat_smooth() is LOESS so we change it to GAM and specify the formula.  (Note that without the "formula = y ~ s(x)" the default for stat_smooth(), even with method = "gam", is to fit a linear regression.)

```{r}
qplot(y = mean,
        x = (julian),
          geom = c("point"),
          data = par.mean) +
    geom_errorbar((aes(ymin = mean-2*SE, 
                       ymax = mean+2*SE)))  +
  stat_smooth(method = "gam", formula = y ~ s(x), size = 1)
```

For more information on stat_smooth() see https://stats.idre.ucla.edu/r/faq/how-can-i-explore-different-smooths-in-ggplot2/

## Model Summaried Data w/GAM

I use a Generalized Additive Model (GAM; Wood 2006) to model how PAR decreases with time. This could be fanicer by adding autocorrelation (eg AR1) to the errors but I don't think that accomplished anything really.

Also, instead of means for 10 AM to 2 PM we could directly model the raw data with random effects with a GAMM (Generalized Additive Mixed Model), but that wouldn't add anything to the analysis.



### Run the GAM

The notation `s(julian)` tells the model to build a smoother for the covariate "julian".

```{r}
gam1 <- gam(mean ~ s(julian), data = par.mean)

```

The `mgcv` library has some basic plotting functionality using the `plot()` function.  See `?plot.gam` for the full details.  The band around the fitted line is the standard error (SE).

```{r}
plot(gam1)
```


### Extract predicted values from fitted GAM

Just like for a regular regression model fitted using the lm() function we can get predictions from a model fit with gam(). These predictions essentially describe the fitted line we plot.  See ?predict.gam for details, though note that we can just call the predict() function, not predict.gam() to get the prediction.

```{r}
preds.out <- predict(gam1)
```


I'll add these to my original dataframe
```{r}
par.mean$preds <- preds.out
```


## Differentiating the smooths in a model (with CIs for derivatives)

I can't remember whether I discovered that you could calculate derivatives for GAMs by reading the help file, then discovered this guy's blog, or found the blog first (I'd prefer the former).

From the Bottom of The Heap Blog by Gavin Simpson who ask the questions "How do we identify where in the series the data are changing?"

Original blog posted:
https://www.fromthebottomoftheheap.net/2014/06/16/simultaneous-confidence-intervals-for-derivatives/

Update on the simultaneous CI issue
https://www.fromthebottomoftheheap.net/2016/12/15/simultaneous-interval-revisited/

And on the GAM issue
https://www.fromthebottomoftheheap.net/2017/03/21/simultaneous-intervals-for-derivatives-of-smooths/

He now has a function to do this
https://gist.github.com/gavinsimpson/e73f011fdaaab4bb5a30


The function can be accessed by downloading the `tsgam` package from GitHub, or by copying and pasting the function directly into an R file.

```{r}
devtools::install_github("gavinsimpson/tsgam")

```


The function is called `Deriv` and is also in this packages `\R` folder.
```{r}
#https://gist.github.com/gavinsimpson/e73f011fdaaab4bb5a30
source("./R/Deriv.R")
```

The derivative function has these key arguments (with their defaults):

* mod: the model
* eps = 1e-7: "the value of the finite difference used to approximate the first derivative."
* newdata: new dataframe over which to make predictions; the function can also build a dataframe for prediction on the fly.  The size of this prediction dataframe is determien by the argument n (see below)
* n = 200: "if newdata is missing the original data can be reconstructed from model and then n controls the number of values over the range of each covariate with which to populate newdata"
* term: the focal term in the model, eg "s(x)"

See also
https://stackoverflow.com/questions/14207250/determining-derivatives-from-gam-smooth-object



```{r}
deriv.from.fderiv <- Deriv(gam1)
```

He's linked his confidence interval functionality directly to R's `confint()` function

```{r}
CIs.from.confint <- confint(deriv.from.fderiv, type = "simultaneous")
```


```{r}
out <- cbind(-1*deriv.from.fderiv$eval,
             CIs.from.confint)
```


```{r}
ggplot(data = out,
       aes(x = julian, y = est)) +
    geom_hline(aes(yintercept = 0)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line() +
    scale_x_reverse() +
    labs(y = "First derivative", x = "Julian")
```






Calculate derivative of the GAM smooth

* see `?predict.gam` for original code 
* the following code notes are copied from there with some additional notes by NLB

"evaluate derivatives of smooths with associated standard errors, by finite differencing..."

## Set up values to make predictions accross

When making predictions using a fitted model the default is to use the original dataframe.  This 

```{r}
x.mesh <- seq(min(par.mean$julian),
              max(par.mean$julian),
              length=dim(par.mean)[1]) ## where to evaluate derivatives
```


```{r}
newd <- data.frame(julian = x.mesh)
X0 <- predict(gam1,newd,type="lpmatrix") 
```




Set up value to perturb data by to approx the differentiation "finite difference interval to approximate differenatiation"
```{r}
eps <- 1e-7 
```


Shift the evaluation mesh
```{r}
x.mesh <- x.mesh + eps 
```


make new data w/shifted values

key to this method is the use of type="lpmatrix" which outputs the coefficients for the underlying GAM equations instead of actual predictions
```{r}
newd <- data.frame(julian = x.mesh)
X1 <- predict(gam1,newd,type="lpmatrix")
```



Look at output of predication - note that there are 9 columns produced, not that 1, b/c the 9 underlying game equations are being called up
```{r}
temp <- round(X1,2)
colnames(temp) <- gsub("julian","j",colnames(temp))
colnames(temp) <- gsub("Intercept","In",colnames(temp))
```


```{r}
head(temp)
```





Divide perturbed and original coefficients 
```{r}
Xp <- (X1-X0)/eps ## "maps coefficients to (fd approx.) derivatives" (?)
```



### Carry out differentiation

The value "9" is used b/c that "basis" for the GAM has 9 terms
```{r}
summary(gam1)
```


```{r}
Xi <- Xp*0 #empty dataframe
```



#Xi[,(i-1)*9+1:9+1] <- Xp[,(i-1)*9+1:9+1] ## Xi%*%coef(b) = smooth deriv i
```{r}
Xi[,1:9+1] <- Xp[,1:9+1] ## Xi%*%coef(b) = smooth deriv i
df <- Xi%*%coef(gam1)                ##  smooth derivative 
df.sd <- rowSums(Xi%*%gam1$Vp*Xi)^.5 ## cheap diag(Xi%*%b$Vp%*%t(Xi))^.5
```



### Plot derivative & CI for line ###

#### Main plot
```{r}
plot(x.mesh,df,
     type="l",
     ylim=range(c(df+2*df.sd,df-2*df.sd)),
     xlab = "Julian Date",
     ylab = "Derivative")
```


#### Plot 95%CI of derivative
```{r}
lines(x.mesh,df+2*df.sd,lty=2);lines(x.mesh,df-2*df.sd,lty=2)
```

```{r}
leg <- c("GAM\nderivative\n",
         "95% CI",
         "Deriv = 0",
         "1st day\nDeriv = 0",
         "1st day\n95% CI = 0")

lty.x <- 2
col.x <- 3

legend("topright",bty = "n",#cex = 0.9,
       legend = leg,
       cex = 0.9,
       lty = c(1,2,1,lty.x,lty.x +1),
       col = c(1,1,2,col.x,col.x+1),inset=c(-0.385,0))
```



```{r}
abline(h = 0, col = 2)

```


## plot deriv = 0 horizontal line 

## plot 1st day when deriv = 0
```{r}
day.deriv.0 <- x.mesh[which(df> 0)[1]]
abline(v = day.deriv.0, col =col.x, lty = lty.x)

text(x = day.deriv.0,y = -30,
     col = 3,
     labels = round(day.deriv.0,0),
     pos = 4)
```


```{r}
date1 <-paste(par1[which(par1$julian == 153)[1], c("month","day")],
              sep = "/",
              collapse = "/")

date1 <- paste(date1,"2016",
               sep = "/")
text(x = day.deriv.0,y = -35,
     col = 3,
     labels = date1,
     pos = 4)

```




## plot 1st day when deriv not sig different from zero

```{r}
day.deriv.0b <- x.mesh[which((df+2*df.sd)> 0)[1]]
abline(v = day.deriv.0b, col = col.x+1, lty = lty.x +1)
text(x = day.deriv.0b-10,y = -30,
     col = 4,
     labels = round(day.deriv.0b,0),pos=4)
date2 <- paste(par1[which(par1$julian == 143)[1], c("month","day")],
               collapse = "/")
date2 <- paste(date2,"2016",sep = "/",
               collapse = "/")
text(x = day.deriv.0b-20,y = -35,
     col = 4,
     labels = date2,pos=4)
```



## References

Wood, S.N.  2006.  Generalized Additive Models: An Introduction with R. Chapman and Hall/CRC.






